---
title: "Extracting model fit statistics in R: AIC and R^2"
draft: false
format:
  html:
    grid:
      margin-width: 400px
reference-location: margin
citation-location: margin
---

# What's in this tutorial?

In the [Unit 4 reading](errors_reading.qmd), we discussed a few different ways to examine and compare models. That is after learning to fit a linear model in [Unit 3](regression_foundations_reading.qmd), Unit 4 discusses some tools for evaluating whether a model is good or not, and whether one model is better than another.

This tutorial explains how to use R to get the quantities and make the comparisons discussed in the Unit 4 reading.

I recommend going through the [Code Tutorial on residuals](R_residuals.qmd) and the [Code Tutorial on model simulations](R_modelsimulations.qmd) as well as the [Unit 4 reading](errors_reading.qmd) before going through this tutorial.


# Loading data and packages

We actually don't need any additional packages in this notebook, but if you want more options and detailed functions related to information-theoretic statistics like AIC (including the "corrected" AIC for small samples), you should check out the `AICcmodavg` package:

[https://cran.r-project.org/web/packages/AICcmodavg/index.html](https://cran.r-project.org/web/packages/AICcmodavg/index.html)


## Load the Pearson & Lee (1903) data

Use the code here to read in a subset of the data from Pearson & Lee (1903), which is a classic data set of the heights of fathers and their sons collected in a survey. This data is derived from the `PearsonLee` data set found in the `HistData` package, and more details can be found in the documentation for that package.

I'm naming the variable "pl" to stand for Pearson & Lee.

```{r}
pl <- read.csv("FatherSonHeights.csv")
head(pl)
```

This data should have 4,312 rows and three variables, though the `observation` variable is just an arbitrary number matching the row number.

The `father_height` and `son_height` variables are expressed as inches. All of the measurements were rounded to the nearest inch but then recorded on the "half-inch". In other words, there are values 62.5, 63.5, 64.5, and so on, but no values 62.0, or any value between the XX.5 values.


# Why use model fit statistics?

There are lots of ways to examine and explore a model. Looking at parameter estimates, residuals, predictions, and simulated data can be good ways to examine the details, but sometimes you might want an "overall" indicator of how well the model is fitting.

Here I describe how to get two of the most common, $R^2$ (aka "R-squared") and AIC (the Akaike Information Criterion).


# Fitting the model

Of course, before we can get any model fit statistics, we need a model!  So we will just fit the same simple model we have before, using the `father_height` variable to predict the `son_height` variable:

```{r}
height_model <- lm(son_height ~ 1 + father_height, data = pl)
```


# Getting and using AIC

Getting the AIC of a model is "stupid simple", because there's an `AIC` function (again, just remember that capitalization matters in R.)

```{r}
height_model_aic <- AIC(height_model)
print(height_model_aic)
```

Using AIC is a little more nuanced. In short:

- The lower the number is, the better of a model it is.
- The value of the number itself is pretty meaningless, and depends a lot on the scale of the variables in the model and how many observations you have (more observations usually means bigger number).
- So you can only compare models that have the *EXACT SAME $y$ value*.

For example, if you had a model that was fitted to some data, and then you excluded some observations and re-fit the model, the AIC would change. The same with transformation, if you transform the $y$ variable (the response variable) of the model in any way, the AIC will change. This doesn't mean it's a bad idea to transform your $y$ variable; it just means that you cannot use AIC to compare models where one has a transformed $y$ and the other doesn't. As long as the $y$ variables are *identical*, you can use AIC to compare models.

So just to say it one more time a slightly different way: while you *can* and *should* use AIC to compare models with different *predictors*, you need to make sure that the *response* variable values are identical in order to compare the AIC values.

To see this in action, let's fit a model that uses a random number to predict `son_height` instead of `father_height`. Because this model and our previous model have the exact same $y$ variable (the values we are trying to predict), we can compare their AIC values.

```{r}
pl$random <- rnorm(nrow(pl))
random_model <- lm(son_height ~ 1 + random, data = pl)

AIC(height_model)
AIC(random_model)
```

The AIC from the `height_model` is lower, so it's a better model.

The final caveat is to not consider "how much lower". That is, the difference between the AICs of these two models is only like 1300 out of around 20,000, so it's tempting to think that maybe they aren't so different. But this kind of comparison actually doesn't make any sense with the mathematical logic of AIC, so just avoid the temptation to think about the numbers like that. Lower just means better. AIC is very good as a *comparison* metric.

If you want something that's more of an *absolute* indicator of how well a model is doing, you need something like $R^2$, which we will look at next.


# Getting and using $R^2$

Getting the $R^2$ ("R-squared") value from a model is easy in R. Just like we did with the residual standard error (see the ), we can extract the $R^2$ value from the model summary object.

Let's get the $R^2$ values from both the height model and the "random" model we fit above.  Instead of saving the model summary as an object first (which is what I did in the [Code Tutorial on model simulations](R_modelsimulations.qmd)), I'll just grab it directly. I'm doing this to show you another way of doing things, even though it might look a little odd at first.

```{r}
height_model_rsquared <- summary(height_model)$r.squared
random_model_rsquared <- summary(random_model)$r.squared

height_model_rsquared
random_model_rsquared
```

So the $R^2$ of our height model could be rounded off to 0.264, which can be interpreted as saying that using the fathers' heights explains around 26% of the variation seen in the heights of their sons. The $R^2$ of the random model will change every time we run the code since we are generating random numbers, but it will be very small, less than 0.001. (See the side bar note for reading scientific notation, which R will tend to use for very small or large numbers.) by convention, we usually only pay attention to two or three digits of $R^2$ values, so we would report these results as follows:

::: column-margin
Note that for the `random_model`, the $R^2$ may be so small that it uses scientific notation. In R, this may look something like `1.2345e-05`. The number after the `e` tells you how many places to move the decimal point, and the "-" or "+" tells you which direction, where the "-" means "make the number smaller by moving the decimal to the left." For example, `1.2345e-05` could also be written as `0.000012345`, but `1.2345e+10` could be written `12345000000`.
:::

- The model using father height to predict son height has an $R^2$ = .264
- The model using a random number to predict son height has an $R^2$ < .001


Interpreting $R^2$ is a little more straightforward than the more "abstract" AIC values. So here, we can say something like our height model is "explaining" roughly a fourth (a little over .25) of the variation in son heights. That's pretty good, and nothing to sneeze at, but clearly there's a lot more variation still to explain. Whereas in the random model, it's explaining essentially nothing, since the $R^2$ is close to zero. This makes sense, because if a completely random number was a good predictor, it would be pure luck, like trying to predict how long you'll live by rolling a bunch of dice at random.


# Summary

While it's *always* important to look closely at how your model is performing, by looking at detailed information such as residuals, it's also sometimes useful to have a single number that represents an "overall" assessment of how good the model is, that you can use to make rough comparisons with other models.

AIC is a very useful and robust statistic that can compare models, as long as they are modeling the exact same $y$ values, but it doesn't have any direct interpretation that is useful. In short, don't worry about the values, just check for which one is lower.

On the other hand, the $R^2$ statistic has a very intuitive interpretation, and it can give you an overall sense of how "effective" your model is at explaining the variance in the $y$ variable. One of the downsides of $R^2$ is that once you get beyond simple linear models, it becomes harder or even impossible to calculate, while AIC applies to a very wide range of model types. 

In the end, both of these measures has pros and cons, but they both have their uses. Fortunately, they are both very easy to extract/compute in R after you have fit your model.

