---
title: "INST 314 week 2: Univariate Distributions"
author: "Dr. Scott Jackson"
date: "2024-06-05"
output:
  tufte::tufte_handout: default
  tufte::tufte_html: default
---

```{r, echo=FALSE}
library(tufte)
```

# What's in this document?

This is the primary reading for the second unit of the course. It was written using R Markdown, so I am providing both the (easier-to-read) PDF rendering as well as the R Markdown "source", if you find that to be useful.

The overall goal of this reading is to continue the journey of exploratory data analysis by looking more deeply at *univariate* distributions and analysis, including important theoretical concepts like the Central Limit Theorem.

I will call attention to key terminology with margin notes. I will also ramble about side thoughts in margin notes as well. Sometimes the margin notes will be numbered,[^sidenote] like footnotes that connect to a specific part of the main text.  Sometimes they're unnumbered, just hanging out whenever you feel like reading them.

```{marginfigure}
This is a margin note.
```

[^sidenote]: This is a numbered side-note.

R code illustrating the concepts here will be explained in more detail in other documents.

# Taking EDA further

To recap a bit, we ended up last unit beginning the process of *exploratory data analyis* (EDA) by examining some basic aspects of *variables* in a *data frame*.[^readings] For categorical variables (including *ordinal* and *nominal* variables), we wanted to get a tabulation of counts of the different values, and for numerical values (*continuous* or *interval* variables in particular), we wanted to get some numerical summaries like the range (min/max) and and central (mean/median) vales.

[^readings]: General concepts from the `INST314_reading_databasics` file, R details in Notbooks on fundamentals, vectors, and data frames.

But recall that the overall goals of EDA are to check and challenge assumptions and expectations. That is, we want to make sure that there isn't anything wrong with the data by confirming some basic assumptions, but we also want to look for surprising results that we may not have expected, because that may shape our analysis to come.

In order to sharpen what we mean by "assumptions", "expectations," and "surprise", we need to dive more deeply into *univariate* analysis[^univariate] and flesh out the statistical concepts behind univariate distributions.

[^univariate]: **Univariate**: referring to a single variable at a time.

# Models and parameters

First, let's step back a bit. One of the biggest challenges of *any* science is to discuss phenomena, data, and patterns at the right level of complexity or specificity. Think about medical science. Technically, every human body is unique in at least some ways. Even identical twins -- the closest nature gets to making "clones" -- are slightly different. So if we gather data about a single body, how can we know that what we learn will apply to the next body we examine? Or even more basic, if chemists are able to isolate a molecule of water and verify that it is made up of two hydrogen atoms bonded with an oxygen atom, how do we know that other water molecules have the same make-up?

One way of answering this epistemological challenge is through the concept of *models*. A model is just a simplified way of looking at the world. We constantly construct mental models (consciously and unconsciously) to help ourselved navigate the world, because at the end of the day, the world is just too complex to understand at the finest levels of detail, and we need models to help us approach things in a more manageable way.

In statistics, the term *model* is used in a few different ways, but it always has this core meaning: a simplified expression of some underlying process that we are trying to describe or analyze. When we talk about "assumptions" in statistics, what we usually mean is that we are "assuming" a particular model, and that model is designed to capture certain properties or patterns. In other words, because a model is a simplification, applying a model to a process means that you are (implicitly or explicitly) making a bunch of assumptions about what the data from that process looks like.

The other abstract concept we need to wrap our heads around is the concept of a *parameter*. Most statistical models have something like "settings," where if we tweak the settings of the model, it can describe different patterns. The technical term for one of these "settings" is a *parameter*.[^nonparametric]

[^nonparametric]: There are also *nonparametric* statistics, but we may only touch on those at the end of the course, if at all. They represent methods for analyzing data when we have so much uncertainty that we think that assuming a model with *any* parameters may be not cautious enough.

# Univariate distributions

But this is all very abstract so far. In order to make this a little less abstract, let's talk about a real-world process, and how we can construct a model for it.

Let's talk about rolling dice.

## Rolling dice: a generating process

One useful way to think about your data is to think about the *generating process* of that data. Where did the data come from? What determines the values?

```{marginfigure}
Note that a generating process is not the same as the *data collection* process, though the latter might affect the former. The data collection process is a description of how the data was obtained or recorded. For example, did you call people to ask their opinion on who they will vote for? How did you pick whom to call? Did you make decisions about throwing out some data (for example, if someone gave nonsensical answers)?

In contrast, the data *generating* process is more of a theoretical concept of all of the underlying factors that give rise to the value that was observed. This could include aspects of data collection, especially if the data collection method introduced any bias, but it is often about unobservable things. For example, what makes a person decide who to vote for? What makes a person decide what to *say* when someone asks them who they plan to vote for? 
```

One of the simplest generating processes to think about is flipping a coin or rolling a die, because it's something we can do physically. So imagine the following process:

1. roll a single six-sided (cube) die
2. write down the number on a slip of paper
3. put that slip of paper in a bag
4. repeat steps 1 through 3 many, many times

The result is that you have a bag full of numbers, generated by a process.

## Process, population, and sample

Now we can think about how this analogy fits with the kinds of data we are actually interested in. The three important concepts here are the *generating process*, the *population* of data that exists (or could exist) in the world, and the *sample* of data that we actually observe.

In our simple dice scenario, rolling the die is the *generating process*, all of the numbers in the bag are the *population*, and any numbers we pull out of the bag to examine are the *sample*.

Whenever we collect data, we are taking a sample. Imagine that the world is a huge bag of numbers and any time we make observations or try to analyze some data, we can only pull a few of those numbers from the bag. For example, if we are interested in the average height of 3-year-old males (perhaps because we want to know if our 3-year-old son is on track, growth-wise), the population (like the entire bag) might be all of the 3-year-old males in the world[^popselection], and clearly getting all of that data is not feasible, or even necessary. But we have to collect **some** data in order to have some idea, so we collect a *sample*. It might be a large sample or a small sample, but it's still a sample.

[^popselection]: Now think to yourself, "would I really want to compare my kid to all the kids in the world?" Given that different parts of the world have different ethnic tendencies, different nutritional tendencies, etc., that may not be a good comparison. This is just a taste of what analysts need to think about when they are making comparisons, thinking about what the right population is to generalize to, and so on.

How you get your sample is of course really important. Consider how political polls are done. Pollsters collect data by talking to just a sample of people, and then using that data to make guesses about how the entire voting population might vote in an actual election. But not all samples are the same! Asking the first ten people you see in a fancy restaurant in Chevy Chase is different than asking the first ten people you see sleeping in tents in DC. These are all valid data (assuming all of those people are eligible to vote), but they are very different samples, and they may not give you a good picture of the overall pattern in the population.

But back to the dice analogy, virtually any actual data collection is like pulling a set of numbers from the bag. If you are somehow in a position to be able to get *every* number in the bag, you might be able to get the entire population. But at the end of the day, what we are really interested in is constructing a *model* of the generating process itself. How did the pattern of numbers get to be that way?

## Theoretical vs. empirical distributions

So let's think about rolling a (fair) die. There are six sides, so we have an equal chance of getting a 1, 2, 3, 4, 5, or 6, every time we roll the die. So if we pull 600 numbers from our bag of numbers, we might expect to see 100 occurrences of each number, right? But what if we didn't?

Now think about if we roll the die just 6 times. What do you think the chances are that we would see exactly 1 occurence of each number?

The fact is, there are many, many possible ways for lots of die rolls to go, even if the die is fair. If you need to convince yourself, you can try it, and start rolling sequences of 6 rolls at a time, and look at how many different sequences you get.

In other words, our sample is rarely a perfect reflection of the underlying process, even when that process is simple. This is where we come to the concept of *empirical* vs. *theoretical* distributions.

First, a *distribution* is just the term we use to talk about what values are more or less likely or common. So when we look at the numbers we pull from the bag and see how many we have of each, that's examining the distribution of values. A *theoretical* distribution is the distribution we expect from a particular model. So if our model is "each side of the die is equally likely," then our theoretical distribution if we looked at a sample of 600 numbers would be 100 of each number. In contrast, the numbers we actually observe are the *empirical* distribution.

In other words, the theoretical distribution describes the generating process[^theoreticapop], and the empirical distribution describes our data sample.

[^theoreticapop]: By extension, the theoretical distribution will also typically describe the population we are trying to generalize to.

Both of these are important when we are doing EDA. We want to examine the empirical distribution, because we want to see what the pattern is for the data we actually have. But then we want to think about how those data compare to theoretical distributions, because we want to check our assumptions of the underlying generating processes of the data.

## The Uniform distribution

Fortunately for us, statisticians have developed a wide range of different mathematical models to capture different theoretical distributions. So our task is often a matter of choosing the distribution that best fits our data.

One of the simplest distributions is called the *uniform* distribution. Conceptually, this is a good choice for our dice example[^uniform], because the idea is that any value has an equal chance of being generated (i.e., the chance of any particular value being generated is "uniform").

[^uniform]: Now technically, this isn't quite accurate because the true uniform distribution is a distribution over *continuous* numbers, and the six values of a die are not (technically) continuous. But it's a good approximation conceptually, just to have a better intuition of how the uniform distribution works.

The uniform distribution is also often called a "flat" distribution, because when we visualize it, it looks flat because all of the values have the same chance of occurring.[^notebook]

[^notebook]: We will actually go through the process of visualization in the accompanying R Notebook.


So now we have a model of the generating process: the results we get from rolling a die are like the results we get from a uniform distribution. But when we talked about models above, we said that models have parameters, like settings. So what are the parameters of the uniform distribution? Take a minute to think about it: if we have a model that says "all values are equally likely", what's the minimum information we need in order to generate numbers that look like rolls from a single die?

The answer is simply the *range*. In other words, the uniform distribution only needs to know what values to pick from, defined by the upper and lower limits. This means that the uniform distribution just has two parameters: the *minimum* value that can be selected, and the *maximum* value that can be selected. The rest of the math behind the uniform distribution is the part that selects any (continuous) value between that minimum and maximum at an equal probability.

# Beyond the Uniform: the Normal distribution

This is fine if our generating processes are all as simple as a die roll, but what about the real world? Most interesting processes are the result of many different factors, combining in many different ways, resulting in more complex distributions.  In other words, many *observed* variables (that is, data you actually collected) may in fact be the result of the combination of many *underlying* or *latent* (unobserved, possibly unobservable) variables.

Fortunately, we have a mathematical rationale that explains why many cases of observed data follow something called the *normal distribution*. 

```{marginfigure}
Note that when we talk about distributions, the word "normal" doesn't have the meaning of "common, typical."  It's a technical term to describe a specific family of theoretical distributions. It's a little confusing because the normal distribution is also very common! Just keep in mind that when we're talking about distributions, the word "normal" is a technical term.
```

## Central Limit Theorem

There is a very powerful and useful mathematical concept called the Central Limit Theorem (CLT). We will skip the math (though the math is very interesting, if you're a math person!) and jump to the core of the concept.

What the theorem says is that when many different (and independent) processes combine (average or sum together), the distribution of that average (or sum) tends to follow a distribution called the *normal* distribution.[^gauss] A simple example of this is that as we've seen, the distribution from a single die is uniform. But if we roll many dice at once, and take the *average* (or sum) of those dice, that average will follow a normal distribution.[^action] 

[^gauss]: Also called the *Gaussian* distribution.

[^action]: Again, we'll see in the in action in the accompanying R Notebook.

That's it! It's ultimately a simple concept -- though perhaps deceptively simple -- but the result is that many of the distributions we end up caring about tend to be normal or close to normal distributions.

## The shape of the normal

In contrast to the "flat" uniform distribution, the normal distribution is sometimes called a "bell curve" because of its shape. But what does that shape mean, and how does it get the shape? By now, you should have at least initial answers for these questions.

```{r, echo = FALSE, fig.margin = TRUE, fig.width = 2.5, fig.height = 2.5, fig.cap = "this is a normal distribution"}
library(ggplot2)

x <- seq(-3, 3, .01)
y_dens <- dnorm(x)
plot_data <- data.frame(x, y_dens)

ggplot(plot_data, aes(x, y_dens)) + geom_line() + theme_void() +
	geom_vline(aes(xintercept = 0), linetype = 2) +
	annotate("text", x = 0.4, y = 0.1, label = "mu", parse = TRUE) +
	geom_segment(x = 0, y = dnorm(1), xend = 1, 
		linetype = 1, arrow = arrow(ends="both", length = unit(0.07, "inches"))) +
	annotate("text", x = 1.35, y = dnorm(1), label = "sigma", parse = TRUE)
```


What the shape of a distribution "means" is that it tells us what values are more or less likely to occur than other values. So the "bell curve" that has a peak ("mode") in the middle that fades to low "tails" in the edges means that values around the peak of the distribution are the most common, and the values at the edges can become exceedingly rare. The precise shape of the distribution gives a precise quantification of "how common" or "how rare."

Where the shape "comes from" is the mathematics of the model, plus the values of its parameters. In other words, the overall pattern of shapes is determined by the theoretical mathematical model, but the specific details of the shape can be altered/adjusted based on its parameters.

The normal distribution only has two parameters:

1. The *mean*:
    - the "central" or "average" value
    - sometimes symbolized with the Greek letter "mu": $\mu$
2. The *standard deviation*: 
   - how "wide" the distribution is, indicating how much values tend to differ from the mean
   - sometimes symbolized with the Greek letter "sigma": $\sigma$

And that's it! With just two parameters you can describe the theoretical shape of a normal distribution.

There is also a special case of the normal distribution called the "standard" normal, which is simply a normal distribution with a mean of 0 and a standard deviation of 1. We will return to standard normals throughout the course.

## Why is the normal important?

So why do we care so much about the normal distribution?

Remember: the CLT tells us that if a (continuous) variable is the result of the combination (average or sum) of a bunch of different processes, that variable will tend to follow a normal distribution, *no matter what the distribution of the component processes are*.

And also recall: it is very often that numbers we collect from real-world data can be a combination of many different underyling (and often unknown) processes. For example, think about someone's score on a standardized test like the SAT. That overall score is the sum of many different factors, including the score on each item, whether the person was well-rested the night before, whether there were unexpected distractions during the testing session, how much time/effort the person spent in preparation, how much effort the person put into years of schooling prior to the test, how much "natural ability" the person may have inherited from their parents, just plain luck, etc. etc. When thinking about any process that resulted in a data value, it's often easy to imagine many many underlying factors contributing to the final observed value.

Put these ideas together, and you can see why many variables we observe in the real world follow a normal distribution. 

And if they are never quite *exactly* normal, they are often "normal enough" that the normal distribution is a very close approximation. Using a convenient theoretical distribution like the normal distribution can help us simplify the math we use to analyze the data. Back to the concept of a *model*, when we use models, we simplify real-world complexity to make it more manageable to analyze. The mathematics of the normal distribution have a lot of convenient and useful properties to make analysis easier. Therefore, as long as the normal is a "good enough" match to the distribution of our data, the conclusions we are able to draw from a model that assumes a normal distribution are also "good enough."[^goodenough]

[^goodenough]: Figuring out what counts as "good enough" is a common challenge and the reason for a great deal of ongoing statistical research.

In other words, we use the normal distribution often because it occurs often, and because it is useful for many different analyses. It is also a useful reference point for other distributions that are very similar, like the *t* distribution, which we will come to later.

## Modes, skew, and tails

When we discuss distributions, usually the best tool is a visualization[^viz], but sometimes we also use words to describe different aspects of a distributional shape.

[^viz]: Examples and practice for how to visualize distributions are in the R Notebook on univariate distributions.

In general, the *mode* of a distribution is just a word for the "high point" of the distribution, which indicates the (theoretically) most likely value. The uniform distribution doesn't really have a mode (or alternatively, the "mode" is the entire range of the distribution), because it's theoretically flat. The normal distribution's mode is exactly at the same point as the mean.[^median] 

[^median]: The median value in a normal distribution also happens to be at the same value.

Many distributions have a portion of the shape that fades into very low probability, like the outer ends of the normal distribution. These are referred to as the "tails" of a distribution. The tails of many distributions (including the normal) never quite reach zero, but just get closer and closer.

Some distributions (including the normal) are symmetrical, meaning that the upper/lower ends (left/right sides in typical visualizations) are mirror images of each other, in theory. If a distribution looks "tilted" one way or another, we talk about it being "skewed." The direction of the skew ("left-skewed" or "right-skewed") just describes the side where the longest "tail" is. So if you have a distribution that looks sort of like a normal, but it looks like it's leaning a bit to the right, with a long left tail, you might say that it's "left-skewed", or "skewed to the left."

```{marginfigure}
There are technical ways to compute and describe skew and other properties like kurtosis -- how "fat" or "skinny" the "bell" part of the curve is, compared to a normal. But we won't go into those because for most purposes, visual plotting of distributions is far more informative than statistical summaries of skew, kurtosis, etc.
```

```{r, echo = FALSE, fig.margin = FALSE, out.height = "2in", out.width = "2in", fig.cap = "this is a left-skewed distribution", warning = FALSE, fig.fullwidth = TRUE}

set.seed(51)
ggplot(data = NULL, aes(x = rnorm(100000)^0.2)) + geom_density() + theme_void()

```

# Next steps

With these concepts out of the way, you should go on to the R Notebook on univariate distributions, to see how these things work in action with a sample data set.
