---
title: "Scatterplots in R"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# What's in this notebook?

The main goal of this notebook is to present and explain the code for creating scatterplots like those in the Unit 3 reading (Regression Fitting).

As a bonus, this will also show you an example of using the `dplyr` package to compute aggregated values. We will use this aggregated values to create the plot where the size of the "dots" reflects the number of observations.

# Loading data and packages

## The Tidyverse

We will continue to use the `ggplot2` package, but we will also use the `dplyr` package. These packages belong to a set of packages that were designed by influential R developer Hadley Wickham referred to as the "tidyverse." The tidyverse packages do things differently than similar functions in base R, but they have a consistent approach and they are extremely useful and powerful. Using tidyverse packages is almost like a "dialect" of R -- some people almost exclusively use them, other people avoid them, but most people use a mix.

There is also a "meta-package" called `tidyverse`, which is essentially just a package that loads all of the packages in the tidyverse. It's not necessary, it's just a convenience. But let's make use of that convenience. So if you haven't already, run:

```{r}
# uncomment this line if you haven't installed this package yet,
# then comment or delete this line after you successfully install
# install.packages("tidyverse")
```

The convenience is that when you install this package, it installs all of the tidyverse packages, and when you load it with `library`, it loads all of those same packages. So if you want to use more than one of them, it can be handy to just load the whole tidyverse.

```{r}
library(tidyverse) # this loads ggplot2, dplyr, and others
```

The tidyverse packages also have some nice online documentation, which includes tutorials, cheatsheets, etc:

https://www.tidyverse.org/


## Load the Pearson & Lee (1903) data

Use this to read in a subset of the data from Pearson & Lee (1903), which is described in the Unit 3 reading. This data is derived from the `PearsonLee` data set found in the `HistData` package.

I'm naming the variable "pl" to stand for Pearson & Lee.

```{r}
pl <- read.csv("FatherSonHeights.csv")
head(pl)
```

This data should have 4,312 rows and three variables, though the `observation` variable is just an arbitrary number matching the row number.

# Basic scatterplot

The first set is just the basic scatterplot. Since we are using `ggplot2`, starting this plot looks a lot like starting a histogram or density plot, because we still use the `ggplot` function to specify what data frame to pull data from, and what columns we want to plot. But where a histogram only has an `x` dimension, a scatterplot has an `x` and a `y`.

Then, it's just a matter of picking the shape or "geom" that we want to plot. In `ggplot2`, the typical geom to use for scatterplots is `geom_point`.

```{r}
ggplot(data = pl, aes(x = father_height, y = son_height)) + geom_point()
```


And just to illustrate, `x` and `y` are the default arguments of `aes`, so if we wanted to be a little lazier, the following code does the same as above:

```{r}
ggplot(pl, aes(father_height, son_height)) + geom_point()
```

## Jittered scatterplots

In the reading, we discussed overplotting, and how one strategy to get around overplotting is to "jitter" the points by adding small random values to the values we're plotting.  There are a few ways to do this in `ggplot2`. The simplest one is to replace the `geom_point` with `geom_jitter`.

```{r}
ggplot(pl, aes(father_height, son_height)) + geom_jitter()
```

Another way to do the same thing is by manipulating the `position` argument of `geom_point`:

```{r}
ggplot(pl, aes(father_height, son_height)) + geom_point(position = "jitter")
```

## Manipulating transparency

Another useful strategy for dealing with overplotting is to adjust the transparency of the things you're plotting. This can be helpful in lots of different kinds of plots, so it's a general property you can adjust for many different geoms. The property is called `alpha`, and it's more like a measure of how opaque something is, from 0 to 1, like a percentage. So an alpha of 0 means 0% opacity, which just makes it invisible. On the other end, an alpha of 1 means 100% opacity, which is the default.

In practice, I usually start with an alpha around 0.5 (since it's halfway) and then play around with the values from there. In the case of this data, since there's so much overplotting, a lower value is probably a little better. But I encourage you to adjust the alpha value up and down in the code below to see how it changes.

```{r}
ggplot(pl, aes(father_height, son_height)) + geom_jitter(alpha = 0.2)
```

## Side-track: aggregation

Aggregating and manipulating data isn't strictly a graphing topic, but it is extremely useful when you are visualizing data, because if you know how to manipulate your data in different ways, you can set up different visualizations.

In this case, look back at the scatterplots here and think about what it means about our data.  Essentially, if you just looked at the `father_height` and `son_height` variables, then it would look like there are a lot of repeated rows.  For example, there are a lot of rows where the value for `father_height` is 67.5 and the value for `son_height` is 68.5. So if we plot a dot for each of those rows, they are plotted on top of each other, which is the "overplotting" pattern we are trying to address in the plots above.

Now what if we reduced things down so that we just had one row for each of the different combinations of father/son heights?  Well, we would lose a lot of data. What data would we lose exactly? We'd lose the count of *how many* father-son pairs are represented by those values. So what we want is a way to reduce down the number of rows to just the unique combinations of the two height variables, but we also want to keep track of how many observations are represented in each of those combos.

This is what aggregation functions are for, because we want to *reduce* the number of rows, but *aggregate* the count of rows that we're in effect collapsing together. This allows us to retain all the information of the full data set.

Here's how to do this using functions from the `dplyr` package. The first step is to create a "grouped" data frame. This sets up how the aggregation will work. We will group the data frame by both of the height variables. This doesn't actually change the data yet, but what it does is set up the data frame so that if we do anything to that data frame, it will do it based on those groups separately.

The next step is to use the `summarize` function, which does the actual aggregation. (Note that this is a different function from the base R `summary` function!) When we use `summarize`, we are basically creating a new column in the new aggregated data frame. To create that column, we give it a name and a value, and we can use whatever functions, variables, etc. to define that value. In this case, all we want is the count of how many values are being aggregated, and there's a special function `n()` that does that. So we can use the `summarize` function on our grouped data frame to make a `count` column that contains the count of how many values were aggregated. Finally, just to clean up a little, we can `ungroup()` our resulting data frame, just to make sure we don't accidentally do other aggregating later. We can always group it again if we want.

```{r}
pl_grouped <- group_by(pl, father_height, son_height)
pl_counts <- summarize(pl_grouped, count = n())
pl_counts <- ungroup(pl_counts)
head(pl_counts)
```

Looking at the head of the resulting aggregated data frame, we see that in the original, unaggregated data, there were four rows where the `father_height` was 58.5 and the `son_height` was 63.5, eight rows where they were 58.5 and 64.5, and so on. (Note, the aggregated data is by default sorted based on the order of the grouping columns, but the sorting of a data frame generally doesn't matter.)


## Back to plotting: using the `size` argument

So why did we go through the above aggregation? Well now we have a data frame that not only has the x's and y's we want to plot (`father_height` and `son_height`), but we have a new column that tells us how many values are at each of those coordinates. And when you have columns in your data, you can map those to graphical properties in `ggplot`!

In this case, we can use the property of the `size` of the points we're plotting to reflect the count of how many father-son pairs are represented with that point. In the `ggplot2` package, every time we want to map something in our *data* to a *graphical property*, we need to use the `aes` function.  The code looks like this (note: I put the geom_point on a different line, which is fine because the + at the end of the first line is a signal to R that the code continues on the second line. This is a useful trick to make your code easier to read, especially when you have long, complex `ggplot` lines.)

```{r}
ggplot(pl_counts, aes(father_height, son_height)) + 
  geom_point(aes(size = count))
```

You may or may not find this easier to understand than the jittered plots, but it's a good option to know about. The advantage to this one is that it is a very faithful representation of the riginal data, which has a lot of repeated values, but it's a lot easier to see that there's a strong trend in the data compared to our very first plot, which had a ton of imperceptible overplotting. 

# Plotting superimposed lines

## Adding "manual" lines

Next, we might like to plot a regression line. There is a handy function for doing that in one step, but let's examine how to plot lines "manually" first.

For example, let's start by plotting a simple 1-to-1 line, which has a slope of 1 and an intercept of 0. This represents a line where the x and y values are exactly the same, because a slope of 1 means that a change in x results in the same change in y as we move up the line, and if it starts at [0, 0], then the x and y values will always be the same.

The `gplot2` package has a geom called `geom_abline` for this, and we can simply add it as a layer to our existing plot:

```{r}
ggplot(pl_counts, aes(father_height, son_height)) + 
  geom_point(aes(size = count)) +
  geom_abline(intercept = 0, slope = 1)
```

This is good, but a black line over black dots might be a little hard to see, so let's make it a dashed blue line by changing the `color` and `linetype` properties of the line:

```{r}
ggplot(pl_counts, aes(father_height, son_height)) + 
  geom_point(aes(size = count)) +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = 2)
```

Now, if we have performed a regression, we will have the slope and intercept parameters for the best-fit regression line. So to add the regression line for this data, all we need to do is run that analysis and plug in those values to the `geom_abline`.

The code for running a regression is covered in a separate notebook, so I won't explain the code, but just use it to get our slope and intercept values from the regression. (But do note that I'm using the full data, not the aggregated data)

```{r}
pl_regression <- lm(son_height ~ father_height, data = pl)
regression_intercept <- pl_regression$coefficients[1]
regression_slope <- pl_regression$coefficients[2]
```

Now we can use these to add the regression line, which I will make red to help it stand out:

```{r}
ggplot(pl_counts, aes(father_height, son_height)) + 
  geom_point(aes(size = count)) +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = 2) +
  geom_abline(intercept = regression_intercept, 
              slope = regression_slope,
              color = "red")
```

## Using the `geom_smooth` function

Being able to add any line you want is very useful, especially when you know how to get the slopes and intercepts. However, there's also a geom called `geom_smooth` which can essentially calculate the regression line for you (as well as other smoothing options, if you want).

First I'll show you the basic use of this geom, but this is not exactly correct for the current data, which I'll explain next. And I'll leave off our other lines for now. (the `se = FALSE` option removes a shaded confidence interval, since we haven't discussed what that means yet)

```{r}
ggplot(pl_counts, aes(father_height, son_height)) + 
  geom_point(aes(size = count)) +
  geom_smooth(method = "lm", se = FALSE)
```

Now, the issue here is that we're adding this line to the aggregated data, because that's what we're plotting. So the regression result we see isn't quite accurate, because it acts as if we only have one data point at each of the rows. But we know that the points in the middle have a lot more data than the outer points, and this make a difference in the regression result.

Luckily, we can have our cake and eat it, too, because we can use a `weights` option to essentially have our `count` variable weight those values in the regression, which gives us the same result as if we had done the regression on the unaggregated data. I will plot the "wrong" regression line in red, and the corrected weighted regression line in blue:

```{r}
ggplot(pl_counts, aes(father_height, son_height)) + 
  geom_point(aes(size = count)) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_smooth(aes(weight = count), method = "lm", se = FALSE, color = "blue")
```

It turns out not to be a huge difference in this example, but you can see that the blue line is a little less steep than the red line. The point here is that if you want to plot regression lines, you can always fall back on using `geom_abline` once you get the actual slope and intercept parameters, but if you want to use `ggplot2` to add regression lines, `geom_smooth` can be convenient. You just need to think about it a bit to make sure it's producing the line you meant for it to.


# Other options and tips

You may have noticed that the plots in the reading sometimes look a little different than the plots you create. One of the nice things about the `ggplot2` package is that the default appearance of graphs looks pretty good, but you also have a lot of control over the final appearance if you want to tweak things.

The downside is that it can get *very* fiddly to make certain kinds of adjustments. It's beyond the scope of this course to do a deep-dive into all the various options, but I'll show you a few of the easier things you can do here.

## Themes

First, there's a powerful *theme* system in `ggplot2`, which is a way of sort of compiling various options into a unified theme. Luckily, `ggplot2` comes with a few nice themes built in to choose from.  Just compare the following:

```{r}
ggplot(pl, aes(father_height, son_height)) + geom_jitter(alpha = .1) + 
  theme_gray()
ggplot(pl, aes(father_height, son_height)) + geom_jitter(alpha = .1) + 
  theme_void()
ggplot(pl, aes(father_height, son_height)) + geom_jitter(alpha = .1) + 
  theme_minimal()
ggplot(pl, aes(father_height, son_height)) + geom_jitter(alpha = .1) + 
  theme_classic()
ggplot(pl, aes(father_height, son_height)) + geom_jitter(alpha = .1) + 
  theme_bw()
ggplot(pl, aes(father_height, son_height)) + geom_jitter(alpha = .1) + 
  theme_dark()

```

Sometimes just picking a good theme can go a long way towards making your plots more readable or enjoyable to look at.

More generally, there is a `theme` function that will also allow you to override any of the settings in these built-in themes. If you ever find yourself wanting to tweak any parts of the plot that aren't the data themselves, it's often the case that it's a setting or parameter that you can tweak inside `theme`. I have the following page bookmarked and I frequently refer to it myself whenever I am trying to fiddle with some of the nitty-gritty of a plot:

https://ggplot2.tidyverse.org/reference/theme.html

## Axis and labels and plot title

Another easy thing to do is to adjust the axis labels. Sometimes you find yourself wanting column names that are easy to work with in R (no spaces, lowercase, etc.), but then they are ugly to look at in a plot, because by default, `ggplot2` uses the names of the columns when it labels things like axes, legends, etc.

Fortunately, it's very easy to tweak this. So for example, I like having simple column names like `father_height` and `son_height` in my data, but they are not pretty axis labels. Here's how I can adjust those, and how to add a "title" to the plot:

```{r}
ggplot(pl_counts, aes(father_height, son_height)) + 
  geom_point(aes(size = count)) +
  xlab("father height (inches)") +
  ylab("son height (inches)") +
  ggtitle("Data from Pearson & Lee (1903)")
```

# Summary

Scatterplots are a powerful tool for visualizing the relationship between two numeric (continuous or interval) variables in your data. Review the discussion of Anscombe's Quartet in the Unit 3 reading if you need any more convincing that visualizing your data is not just a "nice-to-have" step in regression analysis -- it's absolutely crucial, and should come before you bother fitting a model, because a good scatterplot may help you realize that you need to make some changes before fitting and interpreting a model!
